---
title: "Brain Transcriptomics in Alzheimer's Disease"
subtitle: "Postmortem bulk RNA-seq of prefrontal cortex and cerebellum from cognitively normal, Alzheimer's disease and Down syndrome with AD donors"
author: "Rahel Park"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    incremental: false
    smaller: true
    scrollable: true
    embed-resources: true
    fontsize: 1.6em
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| cache: false
#| include: false       

library(tidyverse)
library(edgeR)
library(Glimma)
library(biomaRt)
library(dplyr)
library(stringr)
library(RColorBrewer)
library(gplots)
library(knitr)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)


# ============================================================
# LOAD COUNTS
# ============================================================
counts <- read.delim("data/DSAD-ApoE125_readcounts.txt", 
                     header = TRUE, 
                     skip = 1)
head(counts)

# Remove corrupted rows
bad_rows <- which(is.na(counts$Geneid) | !grepl("^ENSG", counts$Geneid) | 
                    rowSums(is.na(counts[, grepl("^R", colnames(counts))])) > 0)
length(bad_rows)
counts_clean <- counts[-bad_rows, ]

# Drop annotation columns, clean sample column names
counts_clean <- counts_clean %>%
  dplyr::select(-Chr, -Start, -End, -Strand, -Length)
colnames(counts_clean) <- gsub("_Aligned.sortedByCoord.out.bam", "", colnames(counts_clean))

count_matrix <- as.matrix(counts_clean[, -1])
rownames(count_matrix) <- counts_clean$Geneid

# ============================================================
# LOAD AND CLEAN METADATA
# ============================================================
meta_raw <- read.csv("data/EMTAB14179_metadata.csv", check.names = FALSE)
colnames(meta_raw)

meta <- meta_raw %>%
  dplyr::select(
    sample_name = `Source Name`,
    age          = `Characteristics[age]`,
    brain_part   = `Characteristics[organism part]`,
    disease      = `Characteristics[disease]`,
    sex          = `Characteristics[sex]`,
    file_name    = `Comment[SUBMITTED_FILE_NAME]`,
    ENA_run      = `Comment[ENA_RUN]`,
    braak        = Braak,
    pmi          = PMI,
    adrc         = ADRC
  ) %>%
  mutate(
    sample_name = str_replace_all(sample_name, " ", "_"),
    brain_part  = str_replace_all(brain_part, " ", "_"),
    disease = case_when(
      disease == "normal" ~ "CTL",
      disease == "Down syndrome with Alzheimer's disease" ~ "DSAD",
      disease == "Alzheimer's disease" ~ "AD",
      TRUE ~ disease
    ),
    braak = as.factor(braak),
    adrc  = as.factor(adrc)
  )

meta <- meta %>%
  mutate(R_number = str_extract(file_name, "R[0-9]+")) %>%
  mutate(R_number = case_when(
    sample_name == "CTL_2_Cbl" ~ "R54_2",
    sample_name == "DSAD_21_Ctx" ~ "R115_2",
    sample_name == "sAD_9_Cbl" ~ "R61_2",
    TRUE ~ R_number
  )) %>%
  mutate(individual_id = str_remove(sample_name, "_Cbl$|_Ctx$")) %>%
  mutate(group_tissue = paste0(disease, "_", 
                               ifelse(brain_part == "prefrontal_cortex", "PC", "Cer"))) %>%
  dplyr::select(-file_name) %>%
  distinct()

# Check alignment with count matrix
setdiff(colnames(count_matrix), meta$R_number)
setdiff(meta$R_number, colnames(count_matrix))

# Reorder meta to match count_matrix column order
meta <- meta[match(colnames(count_matrix), meta$R_number), ]
all(colnames(count_matrix) == meta$R_number)  # should be TRUE

# ============================================================
# BUILD DGEList
# ============================================================
dge <- DGEList(counts = count_matrix)
dge$samples$group        <- meta$disease
dge$samples$brain_part   <- meta$brain_part
dge$samples$sex          <- meta$sex
dge$samples$age          <- meta$age
dge$samples$individual_id <- meta$individual_id
dge$samples$braak        <- meta$braak
dge$samples$pmi          <- meta$pmi
dge$samples$adrc         <- meta$adrc
dge$samples$group_tissue <- as.factor(meta$group_tissue)

head(dge$samples)
dim(dge)

# ============================================================
# GENE ANNOTATION (biomaRt)
# Cached to disk after the first successful query, since Ensembl's
# public BioMart service is occasionally unavailable and this
# annotation doesn't change between renders.
# ============================================================
annotation_cache <- "data/gene_annotation.rds"

if (file.exists(annotation_cache)) {
  genes_bm <- readRDS(annotation_cache)
} else {
  mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
  genes_bm <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype", "description"),
                    filters = "ensembl_gene_id",
                    values = rownames(dge),
                    mart = mart)
  saveRDS(genes_bm, annotation_cache)
}

nrow(genes_bm)
nrow(dge)

genes_bm <- genes_bm[!duplicated(genes_bm$ensembl_gene_id), ]
genes_bm <- genes_bm[match(rownames(dge), genes_bm$ensembl_gene_id), ]

nrow(genes_bm) == nrow(dge)  # should be TRUE
sum(is.na(genes_bm$ensembl_gene_id))  # how many unmatched

dge$genes <- genes_bm
dge


# ============================================================
# FILTER LOW-EXPRESSION GENES
# ============================================================
keep.exprs <- filterByExpr(dge, group = dge$samples$group)
dge <- dge[keep.exprs, , keep.lib.sizes = FALSE]
dim(dge)

# ============================================================
# NORMALIZATION 
# ============================================================
dge_unnorm <- dge
dge_unnorm$samples$norm.factors <- 1

dge <- normLibSizes(dge, method = "TMM")
dge$samples$norm.factors



# ============================================================
# CONVERT KEY VARIABLES TO FACTORS
# ============================================================
dge$samples$group      <- as.factor(dge$samples$group)
dge$samples$brain_part <- as.factor(dge$samples$brain_part)
dge$samples$sex        <- as.factor(dge$samples$sex)
# braak, adrc, group_tissue already factors - confirmed by str()
# pmi, age stay numeric - correct as-is


# ============================================================
# CORTEX SUBSET: AD vs CTL
# (DSAD excluded — different site/age, addressed separately
# in cross-tissue interaction analysis)
# ============================================================
dge_pc <- dge[, dge$samples$brain_part == "prefrontal_cortex" &
                dge$samples$group %in% c("AD", "CTL")]
dge_pc$samples <- droplevels(dge_pc$samples)

keep.exprs.pc <- filterByExpr(dge_pc, group = dge_pc$samples$group)
dge_pc <- dge_pc[keep.exprs.pc, , keep.lib.sizes = FALSE]
dge_pc <- normLibSizes(dge_pc, method = "TMM")


# ============================================================
# CONFOUNDER CHECKS: AD vs CTL cortex
# 
# ============================================================
# Age difference between groups — actually test it
kruskal.test(age ~ group, data = dge_pc$samples)

# PMI difference
kruskal.test(pmi ~ group, data = dge_pc$samples)

# Sex balance — chi-square
chisq.test(table(dge_pc$samples$group, dge_pc$samples$sex))

# ============================================================
# Cortex MDS and DE objects# 
# ============================================================

lcpm_pc <- cpm(dge_pc, log = TRUE)
design_pc <- model.matrix(~0 + group + age + sex, data = dge_pc$samples)
colnames(design_pc) <- gsub("group", "", colnames(design_pc))
contr_pc <- makeContrasts(AD_vs_CTL = AD - CTL, levels = colnames(design_pc))
v_pc <- voom(dge_pc, design_pc, plot = FALSE)
vfit_pc <- lmFit(v_pc, design_pc)
vfit_pc <- contrasts.fit(vfit_pc, contrasts = contr_pc)
efit_pc <- eBayes(vfit_pc)
top10_pc <- topTable(efit_pc, coef = "AD_vs_CTL", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val, description)

# ============================================================
# EXPLORATORY: SEX-STRATIFIED AD vs CTL (cortex)
# ============================================================

# females
dge_pc_f <- dge_pc[, dge_pc$samples$sex == "female"]
dge_pc_f$samples <- droplevels(dge_pc_f$samples)
keep <- filterByExpr(dge_pc_f, group = dge_pc_f$samples$group)
dge_pc_f <- dge_pc_f[keep, , keep.lib.sizes = FALSE]
dge_pc_f <- normLibSizes(dge_pc_f, method = "TMM")
design_f <- model.matrix(~0 + group + age, data = dge_pc_f$samples)
colnames(design_f) <- gsub("group", "", colnames(design_f))
v_f <- voom(dge_pc_f, design_f, plot = FALSE)
vfit_f <- lmFit(v_f, design_f)
vfit_f <- contrasts.fit(vfit_f, makeContrasts(AD_vs_CTL = AD - CTL, levels = colnames(design_f)))
efit_f <- eBayes(vfit_f)
top_f <- topTable(efit_f, coef = "AD_vs_CTL", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val)

# males
dge_pc_m <- dge_pc[, dge_pc$samples$sex == "male"]
dge_pc_m$samples <- droplevels(dge_pc_m$samples)
keep <- filterByExpr(dge_pc_m, group = dge_pc_m$samples$group)
dge_pc_m <- dge_pc_m[keep, , keep.lib.sizes = FALSE]
dge_pc_m <- normLibSizes(dge_pc_m, method = "TMM")
design_m <- model.matrix(~0 + group + age, data = dge_pc_m$samples)
colnames(design_m) <- gsub("group", "", colnames(design_m))
v_m <- voom(dge_pc_m, design_m, plot = FALSE)
vfit_m <- lmFit(v_m, design_m)
vfit_m <- contrasts.fit(vfit_m, makeContrasts(AD_vs_CTL = AD - CTL, levels = colnames(design_m)))
efit_m <- eBayes(vfit_m)
top_m <- topTable(efit_m, coef = "AD_vs_CTL", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val)


# ============================================================
# CEREBELLUM SUBSET: AD vs CTL
# ============================================================
dge_cer <- dge[, dge$samples$brain_part == "cerebellum" & 
                 dge$samples$group %in% c("AD", "CTL")]
dge_cer$samples <- droplevels(dge_cer$samples)

keep.exprs.cer <- filterByExpr(dge_cer, group = dge_cer$samples$group)
dge_cer <- dge_cer[keep.exprs.cer, , keep.lib.sizes = FALSE]
dge_cer <- normLibSizes(dge_cer, method = "TMM")

dim(dge_cer)
table(dge_cer$samples$group)


# ============================================================
# CONFOUNDER CHECKS: AD vs CTL cerebellum
# 
# ============================================================
# Age difference between groups — actually test it
kruskal.test(age ~ group, data = dge_cer$samples)

# PMI difference
kruskal.test(pmi ~ group, data = dge_cer$samples)

# Sex balance — chi-square
chisq.test(table(dge_cer$samples$group, dge_cer$samples$sex))


# ============================================================
# CEREBELLUM MDS: AD vs CTL
# ============================================================
lcpm_cer <- cpm(dge_cer, log = TRUE)
design_cer <- model.matrix(~0 + group + age + sex, data = dge_cer$samples)
colnames(design_cer) <- gsub("group", "", colnames(design_cer))
contr_cer <- makeContrasts(AD_vs_CTL = AD - CTL, levels = colnames(design_cer))
v_cer <- voom(dge_cer, design_cer, plot = FALSE)
vfit_cer <- lmFit(v_cer, design_cer)
vfit_cer <- contrasts.fit(vfit_cer, contrasts = contr_cer)
efit_cer <- eBayes(vfit_cer)
top_cer <- topTable(efit_cer, coef = "AD_vs_CTL", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val, description)
top_pc <- topTable(efit_pc, coef = "AD_vs_CTL", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, logFC, P.Value, adj.P.Val)

# ============================================================
# BRAAK GRADIENT ANALYSIS: CORTEX (CTL + AD, age + sex adjusted)
# ============================================================

# Use CTL + AD cortex samples with non-NA Braak
dge_braak_pc <- dge[, dge$samples$brain_part == "prefrontal_cortex" &
                   dge$samples$group %in% c("AD", "CTL") &
                   !is.na(dge$samples$braak)]
dge_braak_pc$samples <- droplevels(dge_braak_pc$samples)

# Convert braak to numeric for continuous regression
dge_braak_pc$samples$braak_num <- as.numeric(as.character(dge_braak_pc$samples$braak))

dim(dge_braak_pc)
table(dge_braak_pc$samples$braak_num)

keep <- filterByExpr(dge_braak_pc, group = dge_braak_pc$samples$group)
dge_braak_pc <- dge_braak_pc[keep, , keep.lib.sizes = FALSE]
dge_braak_pc <- normLibSizes(dge_braak_pc, method = "TMM")


design_braak_pc <- model.matrix(~braak_num + age + sex, data = dge_braak_pc$samples)
v_braak_pc <- voom(dge_braak_pc, design_braak_pc, plot = FALSE)
vfit_braak_pc <- lmFit(v_braak_pc, design_braak_pc)
efit_braak_pc <- eBayes(vfit_braak_pc)
top_braak_pc <- topTable(efit_braak_pc, coef = "braak_num", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val, description)

# ============================================================
# BRAAK GRADIENT ANALYSIS: CEREBELLUM (CTL + AD, age + sex adjusted)
# ============================================================

# Use CTL + AD cerebellum samples with non-NA Braak
dge_braak_cer <- dge[, dge$samples$brain_part == "cerebellum" &
                   dge$samples$group %in% c("AD", "CTL") &
                   !is.na(dge$samples$braak)]
dge_braak_cer$samples <- droplevels(dge_braak_cer$samples)

# Convert braak to numeric for continuous regression
dge_braak_cer$samples$braak_num <- as.numeric(as.character(dge_braak_cer$samples$braak))

dim(dge_braak_cer)
table(dge_braak_cer$samples$braak_num)

keep <- filterByExpr(dge_braak_cer, group = dge_braak_cer$samples$group)
dge_braak_cer <- dge_braak_cer[keep, , keep.lib.sizes = FALSE]
dge_braak_cer <- normLibSizes(dge_braak_cer, method = "TMM")


design_braak_cer <- model.matrix(~braak_num + age + sex, data = dge_braak_cer$samples)
v_braak_cer <- voom(dge_braak_cer, design_braak_cer, plot = FALSE)
vfit_braak_cer <- lmFit(v_braak_cer, design_braak_cer)
efit_braak_cer <- eBayes(vfit_braak_cer)
top_braak_cer <- topTable(efit_braak_cer, coef = "braak_num", n = 10, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val, description)


# ============================================================
# CROSS-TISSUE INTERACTION: CORTEX vs CEREBELLUM
# (within-individual, all three groups: CTL, AD, DSAD)
# Using cerebellum as internal reference per individual
# Site confound less problematic here: within-person Ctx-Cer 
# comparison is site-internal for each group
# ============================================================

# Use full dataset (both tissues, all groups)
design_int <- model.matrix(~0 + group_tissue, data = dge$samples)
colnames(design_int) <- levels(dge$samples$group_tissue)
v_full <- voom(dge, design_int, plot = FALSE)
corfit <- duplicateCorrelation(v_full, design_int,
                               block = dge$samples$individual_id)
vfit_full <- lmFit(v_full, design_int,
                   block = dge$samples$individual_id,
                   correlation = corfit$consensus)
contr_int <- makeContrasts(
  AD_region_effect   = (AD_PC - AD_Cer) - (CTL_PC - CTL_Cer),
  DSAD_region_effect = (DSAD_PC - DSAD_Cer) - (CTL_PC - CTL_Cer),
  levels = colnames(design_int)
)
vfit_full <- contrasts.fit(vfit_full, contrasts = contr_int)
efit_full <- eBayes(vfit_full)

dsad_results <- topTable(efit_full, coef = "DSAD_region_effect",
                          n = Inf, sort.by = "P")
top_dsad_region <- topTable(efit_full, coef = "DSAD_region_effect",
                             n = 20, sort.by = "P") %>%
  dplyr::select(external_gene_name, gene_biotype, logFC, P.Value, adj.P.Val, description)
top_ad_region <- topTable(efit_full, coef = "AD_region_effect",
                           n = 20, sort.by = "P")$external_gene_name
top_dsad_region_names <- top_dsad_region$external_gene_name

ad_pathway_genes <- c("KIF5A", "TSC2", "PEX6", "RAPGEF4", "MCTP2",
                      "TREM2", "AIF1", "C1QA", "C1QB", "TYROBP",
                      "VGF", "SCG2", "SCG3", "APP", "MAPT", "APOE")


# ============================================================
# PATHWAY ENRICHMENT: DSAD REGION EFFECT
# Using clusterProfiler for GO and KEGG enrichment
# ============================================================


# ============================================================
# PREPARE GENE LISTS
# ============================================================


# Significant genes (FDR < 0.05): 883 down + 662 up
dsad_sig <- dsad_results %>% filter(adj.P.Val < 0.05)
cat("Total significant DSAD region effect genes:", nrow(dsad_sig), "\n")
cat("Up:", sum(dsad_sig$logFC > 0), "Down:", sum(dsad_sig$logFC < 0), "\n")

# For enrichment: need Entrez IDs (clusterProfiler works best with these)
# Map Ensembl → Entrez via org.Hs.eg.db
library(org.Hs.eg.db)

sig_entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                    keys = dsad_sig$ensembl_gene_id[!is.na(dsad_sig$ensembl_gene_id)],
                                    columns = "ENTREZID",
                                    keytype = "ENSEMBL") %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(ENSEMBL, .keep_all = TRUE)

cat("Genes with Entrez ID mapping:", nrow(sig_entrez), "\n")

# Universe: all tested genes (for background in enrichment)
universe_entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                         keys = dsad_results$ensembl_gene_id[!is.na(dsad_results$ensembl_gene_id)],
                                         columns = "ENTREZID",
                                         keytype = "ENSEMBL") %>%
  filter(!is.na(ENTREZID)) %>%
  pull(ENTREZID) %>%
  unique()

cat("Universe size:", length(universe_entrez), "\n")

dt_full <- decideTests(efit_full)
top50_genes <- dsad_results$ensembl_gene_id[1:50]
top50_genes <- top50_genes[!is.na(top50_genes)]
lcpm_full <- cpm(dge, log = TRUE)
i <- which(rownames(lcpm_full) %in% top50_genes)
group_tissue_labels <- dge$samples$group_tissue
mycol <- colorpanel(1000, "blue", "white", "red")
gene_labels <- ifelse(
  is.na(dge$genes$external_gene_name[i]) | dge$genes$external_gene_name[i] == "",
  rownames(lcpm_full)[i],
  dge$genes$external_gene_name[i]
)

# ============================================================
# BUILD RANKED GENE LIST FOR GSEA
# ============================================================

# Derive from dsad_results — add rank score
dsad_ranked <- dsad_results %>%
  filter(!is.na(ensembl_gene_id)) %>%
  mutate(rank_score = sign(logFC) * -log10(P.Value))

# Map Ensembl → Entrez
ranked_entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                       keys = dsad_ranked$ensembl_gene_id,
                                       columns = "ENTREZID",
                                       keytype = "ENSEMBL") %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(ENSEMBL, .keep_all = TRUE)

# Join rank scores with Entrez IDs
ranked_genes <- dsad_ranked %>%
  left_join(ranked_entrez, by = c("ensembl_gene_id" = "ENSEMBL")) %>%
  filter(!is.na(ENTREZID)) %>%
  arrange(desc(rank_score))

# Remove duplicate Entrez IDs — keep highest absolute rank score per ID
ranked_genes_dedup <- ranked_genes %>%
  group_by(ENTREZID) %>%
  slice_max(abs(rank_score), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(rank_score))

# Named numeric vector — required format for gseGO/gseKEGG
gene_rank_vec <- setNames(ranked_genes_dedup$rank_score, 
                          ranked_genes_dedup$ENTREZID)

# Verify
cat("Total genes in ranked list:", length(gene_rank_vec), "\n")
cat("Unique Entrez IDs:", length(unique(names(gene_rank_vec))), "\n")
head(gene_rank_vec)
tail(gene_rank_vec)

gsea_kegg <- readRDS("data/gsea_kegg.rds")
# gseKEGG run once and saved — tests enrichment against KEGG pathway database
# (curated metabolic/signaling pathways, e.g. "Alzheimer disease", "Lysosome")

#gsea_kegg <- gseKEGG(
#  geneList      = gene_rank_vec,
#  organism      = "hsa",
#  minGSSize     = 10,
#  maxGSSize     = 500,
#  pAdjustMethod = "BH",
#  pvalueCutoff  = 0.05,
#  verbose       = FALSE,
#  seed          = TRUE
#)

cat("GSEA KEGG significant pathways:", nrow(as.data.frame(gsea_kegg)), "\n")

as.data.frame(gsea_kegg) %>%
  dplyr::select(Description, NES, p.adjust, setSize) %>%
  arrange(p.adjust) %>%
  head(20)

# gseGO run once and saved — tests enrichment against Gene Ontology Biological Process (GO:BP)
# (finer-grained biological processes, e.g. "oxidative phosphorylation", "vesicle trafficking")
# Complementary to KEGG: GO:BP gives process-level detail, KEGG gives pathway-level context


gsea_go   <- readRDS("data/gsea_go.rds")

#gsea_go <- gseGO(
#  geneList      = gene_rank_vec,
#  OrgDb         = org.Hs.eg.db,
#  ont           = "BP",
#  minGSSize     = 10,
#  maxGSSize     = 500,
#  pAdjustMethod = "BH",
#  pvalueCutoff  = 0.05,
#  verbose       = FALSE,
#  seed          = TRUE
#)

cat("GSEA GO BP significant terms:", nrow(as.data.frame(gsea_go)), "\n")

as.data.frame(gsea_go) %>%
  dplyr::select(Description, NES, p.adjust, setSize) %>%
  arrange(p.adjust) %>%
  head(20)



```


## Background & Dataset

Bulk RNA-seq of postmortem AD brain shows **limited separation from
normal aging**: immune and synaptic signals exist but are
heterogeneous and sample-size sensitive. DSAD (trisomy 21 → APP
triplication) shares core AD programs but adds chromosomal dosage
biology, making it a partial rather than complete AD proxy.

The **paired prefrontal cortex-cerebellum design** is underexplored as an
analytical strategy. Cerebellum is relatively spared in AD but not
transcriptionally silent, so within-person regional comparisons can be
more sensitive than single-region group differences.

**Dataset: E-MTAB-14179**, Thorwald et al. 2025 (*Alzheimer's & Dementia*)<br>
Primary paper focused on iron metabolism and ferroptosis at
the protein level; transcriptomic data was deposited but not fully analyzed

42 samples from 23 donors · CTL (n=8) · sporadic AD (n=7) · DSAD (n=8)<br>
19 donors with paired prefrontal cortex + cerebellum · all APOE3/3<br>
Braak, PMI, sex, age available · publicly deposited counts matrix

---

## Pipeline Overview

**Original authors (Thorwald et al. 2025):**  
FASTQ → STAR alignment → Count matrix deposited at E-MTAB-14179


**This reanalysis (from deposited count matrix):**  
edgeR/limma-voom → clusterProfiler GSEA

---

## Sample Characteristics & Analytical Decisions

```{r}
#| label: confounders

# One row per donor. Subsetting to cortex would drop the two donors who
# contributed cerebellum only, so deduplicate on donor instead.
paired_ids <- names(which(table(meta$individual_id) == 2))

meta %>%
  distinct(individual_id, .keep_all = TRUE) %>%
  group_by(disease) %>%
  summarise(
    n = n(),
    pairs = sum(individual_id %in% paired_ids),
    age_mean_sd = paste0(round(mean(age), 1), " ± ", round(sd(age), 1)),
    sex_f_m = paste0(sum(sex == "female"), "/", sum(sex == "male")),
    braak = paste0(min(as.numeric(as.character(braak)), na.rm=TRUE),
                   "–",
                   max(as.numeric(as.character(braak)), na.rm=TRUE)),
    pmi_mean = round(mean(pmi, na.rm=TRUE), 1),
    site = paste(unique(as.character(adrc)), collapse="/")
  ) %>%
  dplyr::rename(Group=disease, N=n, `Complete pairs`=pairs,
         `Age (mean±SD)`=age_mean_sd,
         `Sex (F/M)`=sex_f_m, `Braak range`=braak,
         `Post-mortem interval (mean hours)`=pmi_mean, Site=site) %>%
  kable()

```


**Covariate decisions:**

- **Sex**: balanced (3-4/group), included as covariate
- **Age**: included in AD vs CTL models (within-group variation). 
  For DSAD: only analyzed in the paired cross-tissue model where 
  individual-level confounds including age cancel within pairs
- **Braak**: collinear with group (CTL=0-1, AD=3-6), so not added
- **PMI**: no group difference between AD and CTL (Kruskal-Wallis 
  p=0.72 prefrontal cortex, p=0.85 cerebellum) and negligible 
  correlation with expression (r=0.12) — not added. 3 samples 
  have missing PMI, adding it would reduce n unnecessarily. 
  Confounded with site in DSAD; cancels within pairs in paired model.
- **Site**: DSAD=UCI, AD/CTL=USC. Fully confounded; reported as a caveat rather than adjusted for.


---

```{r}
#| label: QC - Library Size and Normalization
#| fig-cap: "Before and after TMM normalization"
#| include: false
par(mfrow = c(1, 2))
boxplot(cpm(dge_unnorm, log = TRUE), las = 2, main = "Before TMM")
boxplot(cpm(dge, log = TRUE), las = 2, main = "After TMM")
par(mfrow = c(1, 1))

```

## MDS: Brain Region Dominates

```{r}
#| label: mds-full
lcpm_all <- cpm(dge, log = TRUE)
par(mar = c(5,4,4,8), xpd = FALSE)

col.group_full <- brewer.pal(nlevels(dge$samples$group), "Set1")[dge$samples$group]
pch.brain      <- c(16, 17)[dge$samples$brain_part]

mds_full <- plotMDS(lcpm_all, pch = pch.brain, col = col.group_full, cex = 1.5)
title("All samples: disease and brain region")

par(xpd = TRUE)
xlim <- range(mds_full$x); ylim <- range(mds_full$y)
legend(xlim[2]+0.3, ylim[2], legend = levels(dge$samples$group),
       col = brewer.pal(3,"Set1"), pch = 16, title = "Disease", bty = "n")
legend(xlim[2]+0.3, ylim[2]-2.0, legend = levels(dge$samples$brain_part),
       col = "black", pch = c(16,17), title = "Brain region", bty = "n")
par(mar = c(5,4,4,2), xpd = FALSE)
```
Dim 1 (45% variance) separates cerebellum from prefrontal cortex cleanly; disease groups show no comparable separation.<br>
→ Continuing with analyses split by region; paired design used for cross-tissue comparison.

---

## AD vs CTL: No FDR-Significant Genes in Either Region

```{r}
#| label: cortex-cer-mds
#| fig-width: 11
#| fig-height: 5
par(mfrow = c(1,2), mar = c(4,4,4,2))

# Cortex
group_pc     <- dge_pc$samples$group
col.group_pc <- brewer.pal(nlevels(group_pc), "Set1")[group_pc]
pch.sex_pc   <- c(16,17)[dge_pc$samples$sex]
mds_pc <- plotMDS(lcpm_pc, pch = pch.sex_pc, col = col.group_pc, cex = 1.5)
title("Cortex: AD vs CTL")
text(mds_pc$x, mds_pc$y, labels = dge_pc$samples$age, pos = 2, cex = 0.6)
legend("bottomright", legend = c("AD", "CTL", "female", "male"),
       col = c(brewer.pal(3,"Set1")[1:2], "black", "black"),
       pch = c(16, 16, 16, 17), bty = "n", cex = 0.7)

# Cerebellum
group_cer     <- dge_cer$samples$group
col.group_cer <- brewer.pal(nlevels(group_cer), "Set1")[group_cer]
pch.sex_cer   <- c(16,17)[dge_cer$samples$sex]
mds_cer <- plotMDS(lcpm_cer, pch = pch.sex_cer, col = col.group_cer, cex = 1.5)
title("Cerebellum: AD vs CTL")
text(mds_cer$x, mds_cer$y, labels = dge_cer$samples$age, pos = 2, cex = 0.6)
legend("bottomright", legend = c("AD", "CTL", "female", "male"),
       col = c(brewer.pal(3,"Set1")[1:2], "black", "black"),
       pch = c(16, 16, 16, 17), bty = "n", cex = 0.7)

par(mfrow = c(1,1), mar = c(5,4,4,2))
```

- **0 FDR-significant genes in either region**, age + sex adjusted
  (cortex: 7 AD vs 7 CTL; cerebellum: 6 AD vs 8 CTL)
- Top nominal prefrontal cortex hits: VGF↓, CRH↓, SCG3↓ (neuroendocrine/synaptic)
- Top nominal cerebellum hits: entirely non-overlapping with cortex
- Consistent with known AD/aging transcriptional overlap in bulk tissue
- *DSAD excluded from single-region analyses — site (UCI vs USC) 
  fully confounded with group; addressed separately via paired cross-tissue model*


```{r}
#| label: Cortex - Sex-Stratified AD vs CTL
#| include: false

# side by side tables
cat("--- Females ---\n"); top_f
cat("--- Males ---\n"); top_m

# comparison
cat("Female only:\n")
setdiff(top_f$external_gene_name, top_m$external_gene_name)
cat("Male only:\n")
setdiff(top_m$external_gene_name, top_f$external_gene_name)
cat("Shared:\n")
intersect(top_f$external_gene_name, top_m$external_gene_name)

# - No overlap between female and male top hits
# - Females: TTR↓, HSD11B2↑ — neuroprotective/stress
# - Males: C1QA↑, C1QB↑, LAPTM5↑ — complement/microglial
# - Hypothesis-generating at n=3-4/sex; warrants larger study

```



---

## Braak Gradient in Prefrontal Cortex

Binary AD vs CTL comparison yielded 0 FDR-significant genes — 
Braak stage (0→6) tested as a continuous severity predictor 
to increase power. Age and sex adjusted · CTL + AD · n=14  
*(DSAD excluded: all Braak 6, no severity gradient)*

**Result: still no FDR-significant genes**

Top nominal hits nonetheless consistent with AD vs CTL: 
VGF↓, SCG2↓, SCG3↓ — neuroendocrine/synaptic genes declining 
with pathology severity in prefrontal cortex, not in cerebellum

Convergent evidence across two analytical approaches — 
the signal exists but is below detection threshold at this sample size

```{r}
#| label: braak
#| include: false


# Design: Braak as continuous covariate, adjusted for age and sex
# No group term — Braak replaces it as the continuous severity measure

# Cortex top hits
cat("--- Cortex Braak-associated genes ---\n")
top_braak_pc

# Cerebellum top hits
cat("--- Cerebellum Braak-associated genes ---\n")
top_braak_cer

# Comparison
cat("Cortex only:\n")
setdiff(top_braak_pc$external_gene_name, top_braak_cer$external_gene_name)
cat("Cerebellum only:\n")
setdiff(top_braak_cer$external_gene_name, top_braak_pc$external_gene_name)
cat("Shared:\n")
intersect(top_braak_pc$external_gene_name, top_braak_cer$external_gene_name)


top_braak_pc_50 <- topTable(efit_braak_pc, coef = "braak_num", n = 50, sort.by = "P")
top_braak_cer_50 <- topTable(efit_braak_cer, coef = "braak_num", n = 50, sort.by = "P")

cat("Shared in top 50:\n")
intersect(top_braak_pc_50$external_gene_name, top_braak_cer_50$external_gene_name)

top_braak_pc_100 <- topTable(efit_braak_pc, coef = "braak_num", n = 100, sort.by = "P")
top_braak_cer_100 <- topTable(efit_braak_cer, coef = "braak_num", n = 100, sort.by = "P")

cat("Shared in top 100:\n")
intersect(top_braak_pc_100$external_gene_name, top_braak_cer_100$external_gene_name)
```

---

## Cross-Tissue Interaction: DSAD Regional Signature

Within-person prefrontal cortex − cerebellum difference vs CTL  
*(19 complete pairs: CTL n=7, AD n=6, DSAD n=6)*

| Contrast | Down | Up | Total sig. genes |
|----------|------|-----|-----------------|
| AD region effect | 0 | 2 | 2 |
| DSAD region effect | 883 | 662 | **1545** |

DSAD disrupts the normal cortex-cerebellum expression relationship massively.  
Sporadic AD does not, despite similar Braak scores in the most severe AD cases.

**Top DSAD-specific hits**: KIF5A (axonal transport), TSC2 (mTOR/autophagy),  
PEX6 (peroxisomal biogenesis), RAPGEF4 (vesicle trafficking)  
→ convergent on endosomal-lysosomal-autophagy axis

*Paired design: individual-level confounds (genetics, PMI, site) cancel within pairs*

```{r}
#| label: cross-tissue
#| include: false

# Summary table
summary(decideTests(efit_full))

# Top DSAD region genes
top_dsad_region

# AD-relevant pathway genes check
dsad_results %>%
  filter(external_gene_name %in% ad_pathway_genes) %>%
  dplyr::select(external_gene_name, logFC, adj.P.Val)

# Shared vs DSAD-specific
cat("Shared AD+DSAD regional disruption:\n")
intersect(top_ad_region, top_dsad_region_names)
cat("DSAD-specific:\n")
setdiff(top_dsad_region_names, top_ad_region)



```





```{r}
#| label: DSAD Region Effect - Key Genes
#| include: false
plotMD(efit_full, column = "DSAD_region_effect",
       status = dt_full[, "DSAD_region_effect"],
       main = "DSAD: Cortex-Cerebellum regional disruption",
       xlim = c(-10, 10))

```





```{r}
#| label: DSAD Region Effect - Heatmap
#| include: false
#| fig-width: 12
#| fig-height: 10
heatmap.2(lcpm_full[i, ],
          scale = "row",
          labRow = gene_labels,
          labCol = group_tissue_labels,
          col = mycol,
          trace = "none",
          density.info = "none",
          margin = c(6, 8),
          lhei = c(1.5, 8),
          dendrogram = "column",
          cexRow = 0.6,
          cexCol = 0.7,
          main = "Top 50 DSAD region-effect genes")

```

---

## DSAD Region Effect: Volcano

```{r}


#| label: volcano
#| warning: false

top_extreme <- dsad_results %>%
  filter(adj.P.Val < 0.05, abs(logFC) > 2) %>%
  arrange(P.Value) %>%
  head(20) %>%
  pull(external_gene_name)

dsad_volcano <- dsad_results %>%
  mutate(
    sig = adj.P.Val < 0.05,
    label = ifelse(external_gene_name %in% top_extreme, external_gene_name, "")
  )

ggplot(dsad_volcano, aes(x = logFC, y = -log10(P.Value), color = sig)) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_manual(values = c("grey60", "red3"),
                     labels = c("Not significant", "FDR < 0.05")) +
  geom_text_repel(aes(label = label), size = 2.5, 
                  max.overlaps = 30,
                  box.padding = 0.3,
                  segment.size = 0.2) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(x = "Log2 FC (DSAD Ctx−Cer vs CTL Ctx−Cer)",
       y = "-log10(p-value)", color = "") +
  theme_bw() +
  theme(legend.position = "bottom")


```

---

## DSAD Regional Disruption: Pathway Enrichment

```{r}
#| label: gsea-kegg
# ============================================================
# VISUALIZE KEGG GSEA (17 significant pathways)
# ============================================================
as.data.frame(gsea_kegg) %>%
  arrange(NES) %>%
  mutate(Description = factor(Description, levels = Description)) %>%
  ggplot(aes(x = NES, y = Description, size = setSize, color = p.adjust)) +
  geom_point() +
  scale_color_continuous(low = "red", high = "blue", name = "adj. p-value", trans = "log10") +
  scale_size_continuous(name = "Gene count") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw() +
  labs(x = "Normalized Enrichment Score (NES)", y = "", title = "")

```

*Within-person prefrontal cortex−cerebellum difference vs CTL*  
*NES > 0: upregulated in DSAD · NES < 0: downregulated · Color = adjusted p-value*

---

## DSAD Regional Disruption: Pathway Interpretation

16 upregulated pathways dominated by **ribosomal/mitochondrial stress**  
(Ribosome, Oxidative phosphorylation) and **neurodegeneration-relevant** pathways:

- **Lysosome biogenesis**, matching the KIF5A/TSC2/PEX6 individual gene hits
- **Alzheimer disease pathway**: directly relevant
- **Reactive oxygen species**, echoing Thorwald et al.'s ferroptosis findings at the protein level

1 downregulated: **Phototransduction** (cerebellar signaling genes),
relatively suppressed in DSAD prefrontal cortex

→ Transcriptomic pathway disruption converges on the same oxidative stress  
and proteostasis mechanisms reported at protein level in the companion paper

---

## Connection to Published Analysis

| | Thorwald et al. 2025 | This reanalysis |
|--|--|--|
| CTL vs AD PFC | 10 DEGs (incl. VGF) | 0 FDR-sig (VGF top nominal) |
| CTL vs AD cerebellum | 1374 DEGs | 0 FDR-significant |
| Region overlap | None (all comparisons) | None confirmed (AD vs CTL + Braak gradient) |
| Pathway analysis | IPA (commercial) | clusterProfiler GSEA (open source) |
| Braak gradient | Not done | Region-specific signatures |
| Cross-tissue interaction | Not done | 1545 DSAD-specific genes |

*Discrepancy in CTL vs AD cerebellum (1374 vs 0): Thorwald et al. applied
RUVSeq correction (parameters unreported) and DESeq2 without stated age
adjustment or FDR threshold, so a full comparison isn't possible without
complete methodological details*

---

## Key Conclusions

1. **AD vs CTL**: 0 FDR-significant genes in bulk cortex or cerebellum;
top nominal hits (VGF↓, CRH↓, SCG3↓) are consistent with known
neuroendocrine/synaptic decline in the AD literature

2. **Braak gradient in prefrontal cortex**: top nominal hits (VGF↓, SCG2↓, SCG3↓) 
consistent with AD vs CTL — converging evidence across two analytical approaches 
despite no FDR-significant genes at n=14

3. **DSAD prefrontal cortex−cerebellum disruption**: 1545 FDR-significant 
genes in the paired interaction model, enriched for lysosomal, peroxisomal, 
mitochondrial, and oxidative stress pathways, echoing Thorwald et al.'s 
protein-level findings

4. **DSAD is not transcriptomically equivalent to sporadic AD**: regional disruption is massive
in DSAD and negligible in sporadic AD. The site confound is acknowledged, but the
pattern is consistent with genuine biological divergence

---

## Limitations

- 7-8 donors/group, and only 6 complete pairs per disease group: low power; no FDR-significant genes in AD vs CTL
- DSAD samples come from a different site (UCI vs USC), confounded with group, age, and PMI; unresolvable in this design
- Bulk RNA-seq dilutes cell-type-specific signals, so microglial activation isn't detectable


